---
title: "bikedata"
output:
  github_document:
    toc: true
  #html_document:
  #  toc: true
  #  toc_float: true
  #  theme: flatly
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/mpadge/bikedata.svg)](https://travis-ci.org/mpadge/bikedata) 
[![codecov](https://codecov.io/gh/mpadge/bikedata/branch/master/graph/badge.svg)](https://codecov.io/gh/mpadge/bikedata)
[![Project Status: Concept - Minimal or no implementation has been done yet.](http://www.repostatus.org/badges/0.1.0/concept.svg)](http://www.repostatus.org/#concept)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/bikedata)](http://cran.r-project.org/web/packages/bikedata) 

# bikedata


The `bikedata` package aims to enable ready importing of historical trip data
from all public bicycle hire systems which provide data, and will be expanded on
an ongoing basis as more systems publish open data. Cities and names of
associated public bicycle Systems currently included, along with numbers of
bikes and of docking stations, are:

City | Hire Bicycle System | Number of Bicycles | Number of Docking Stations
--- | --- | --- | ---
London, U.K. | [Santander Cycles](https://tfl.gov.uk/modes/cycling/santander-cycles) | 13,600 | 839
New York City NY, U.S.A. | [citibike](https://www.citibikenyc.com/) | 7,000 | 458
Chicago IL, U.S.A. | [Divvy](https://www.divvybikes.com/) | 5,837 | 576
Washingon DC, U.S.A. | [Capital BikeShare](https://www.capitalbikeshare.com/) | 4,457 | 406
Boston MA, U.S.A. | [Hubway](https://www.thehubway.com/) | 1,461 | 158
Los Angeles CA, U.S.A. | [Metro](https://bikeshare.metro.net/) | 1,000 | 65

These data include the places and times at which all trips start and end. Some
systems provide additional demographic data including years of birth and genders
of cyclists.

------


## 1 Installation

Currently a development version only which can be installed with the following
command.

```{r, eval=FALSE}
devtools::install_github("mpadge/bikedata")
```
```{r usage2, echo=FALSE, message=FALSE}
devtools::load_all (".")
#devtools::load_all (".", recompile=TRUE)
#devtools::document (".")
#goodpractice::gp ("bikedata")
#devtools::check (".")
#testthat::test_dir ("./tests/")
#Rcpp::compileAttributes()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
options(width=120)
```


------


## 2 Usage

```{r, eval=FALSE, echo=FALSE}
library(bikedata)

# current verison
packageVersion("bikedata")
```

Data may downloaded for a particular city and stored in an `SQLite3` database
with the simple command:
```{r, eval=FALSE}
store_bikedata (city = 'nyc', bikedb = 'bikedb')
```
where the `bikedb` parameter provides the name for the database.  Numbers of
trips between each pair of stations ('trip matrices') can then be obtained with:
```{r, eval=FALSE}
tm <- bike_tripmat ('bikedb')
dim (tm); format (sum (tm), big.mark=',')
```
```{r, echo=FALSE}
c (689, 689)
"36,902,025"
```
These trip matrices are the primary objects returned by the `bikedata` package.
Note that the associated databases can be very large (in the above case, almost
37 million individual trips), and extracting these data can take quite some
time.

A summary of all data contained in a given database can be produced as
```{r, eval=FALSE}
bike_summary_stats (bikedb = 'bikedb')
#>    num_trips num_stations          first_trip       last_trip latest_files
#> ny  36902025          689 2014-11-01 00:00:11 2017-01-31 9:59        TRUE
```
The final field, `latest_files`, indicates whether the files in the database are
up to date with the latest published files.

### 2.1 Filtering trips by date

Trip matrices can be constructed for trips filtered by dates and/or times. The
temporal extent of the database is given in the above `bike_summary_stats()`
function, or can be directly viewed with
```{r, eval=FALSE}
bike_datelimits (bikedb = 'bikedb')
```
```{r, echo=FALSE}
res <- c ("2013-07-01 00:00:00", "2017-01-19 08:08:17")
names (res) <- c ("first", "last")
res
```
The `bike_tripmat` function accepts several additional arguments, including
values specifying `start_date`, `end_date`, `start_time`, and `end_time`.  A
trip matrix constructed from trips beginning after a given date with the
`start_date` argument.
```{r, eval=FALSE}
tm <- bike_tripmat ('spdb', start_date="20160101")
format (sum (tm), big.mark=',')
```
```{r, echo=FALSE}
"13,845,655"
```
As expected, that reduces numbers of trips. End dates can also be specified
```{r, eval=FALSE}
tm <- bike_tripmat ('spdb', start_date=20140901, end_date="14/09/03")
format (sum (tm), big.mark=',')
```
```{r, echo=FALSE}
"89,508"
```
Note that dates can be specified in almost any format, as long as the order is
`year-month-day`.

### 2.2 Filtering trips by time of day

Trips can also be selected starting and/or ending at specific times of day.
```{r, eval=FALSE}
tm <- bike_tripmat ('spdb', start_time=0, end_time=1)
format (sum (tm), big.mark=',')
```
```{r, echo=FALSE}
"34,512"
```
Single values are interpeted to specify hours, so the above request returns only
those rides start at or after midnight (`00:00:00`) and finishing prior to and
including 1am (`01:00:00`). Single numeric values of `24` are interpreted as the
end of a day (`23:59:59`).

Dates and times can of course be combined
```{r, eval=FALSE}
tm <- bike_tripmat ('spdb', start_date=20150101, end_date="15,12,31",
                 start_time=6, end_time=10)
format (sum (tm), big.mark=',')
```
```{r, echo=FALSE}
"1,635,974"
```

### 2.3 Filtering trips by day of week

Trips can also extracted for particular days of the week by specifying the
`weekday` argument of `bike_tripmat`. Weekdays can be numeric, start from `1=Sunday`,
or any unambiguous character string.
```{r, eval=FALSE}
tm <- bike_tripmat ('spdb', weekday=7)
format (sum (tm), big.mark=',')
```
```{r, echo=FALSE}
"4,469,250"
```

Several weekdays can also be selected
```{r, eval=FALSE}
tm <- bike_tripmat ('spdb', weekday=c('m', 'Th'))
format (sum (tm), big.mark=',')
```
```{r, echo=FALSE}
"10,923,683"
```

Weekdays can also be specified along with other dates and times
```{r, eval=FALSE}
tm <- bike_tripmat ('spdb', weekday=2:6, start_time=6, end_time=10)
format (sum (tm), big.mark=',')
```
```{r, echo=FALSE}
"6,333,545"
```
