---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/mpadge/bikedata.svg)](https://travis-ci.org/mpadge/bikedata) 
[![Project Status: Concept - Minimal or no implementation has been done yet.](http://www.repostatus.org/badges/0.1.0/concept.svg)](http://www.repostatus.org/#concept)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/bikedata)](http://cran.r-project.org/web/packages/bikedata) 

# bikedata

R package to load data from public bicycle hire systems. Currently a
proof-of-concept that only loads data from the New York City [citibike
scheme](https://www.citibikenyc.com/).


## Installation

```{r, eval=FALSE}
devtools::install_github("mpadge/bikedata")
```
```{r usage2, echo=FALSE, message=FALSE}
devtools::load_all (".")
#devtools::load_all (".", recompile=TRUE)
#devtools::document (".")
#goodpractice::gp ("bikedata")
#devtools::check (".")
#testthat::test_dir ("./tests/")
#Rcpp::compileAttributes()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
options(width=120)
```

## Usage

```{r, eval=FALSE, echo=FALSE}
library(bikedata)

# current verison
packageVersion("bikedata")
```

Data from the NYC citibike system can can be downloaded with
```{r, eval=FALSE}
dl_bikedata (data_dir='/tmp')
```
The resultant files can then be read into an `spatialite` database with
```{r, eval=FALSE}
store_bikedata_spl (data_dir='/tmp', "junk")
```
```{r, eval=FALSE, echo=FALSE}
store_bikedata_spl (data_dir='/data/data/bikes/nyc-temp', 'junk')
```

Numbers of trips between each pair of stations can then be obtained in square
matrix form from
```{r, eval=FALSE}
tmat <- tripmat ("junk")
```
