n--
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/mpadge/bikedata.svg)](https://travis-ci.org/mpadge/bikedata) 
[![Project Status: Concept - Minimal or no implementation has been done yet.](http://www.repostatus.org/badges/0.1.0/concept.svg)](http://www.repostatus.org/#concept)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/bikedata)](http://cran.r-project.org/web/packages/bikedata) 
![downloads](http://cranlogs.r-pkg.org/badges/grand-total/bikedata)

# bikedata

R package to load data from public bicycle hire systems. Currently a
proof-of-concept that only loads data from the New York City [citibike
scheme](https://www.citibikenyc.com/).


## Installation

```{r, eval=FALSE}
devtools::install_github("mpadge/bikedata")
```
```{r usage2, echo=FALSE}
devtools::load_all (".")
devtools::load_all (".", recompile=TRUE)
devtools::document (".")
#goodpractice::gp ("bikedata")
#devtools::check (".")
#testthat::test_dir ("./tests/")
#Rcpp::compileAttributes()
```
```{r}
read_bikedata ()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
options(width=120)
```

## Usage

```{r, eval=FALSE}
library(bikedata)

# current verison
packageVersion("bikedata")
```

### Rcpp

```{r}
read_bikedata ()
```

### postgres

[citibike data](https://www.citibikenyc.com/system-data) first have to be
downloaded:
```{r, eval=FALSE}
dl_bikedata ()
```
And stored in a `postgres` database with
```{r, eval=FALSE}
store_bikedata (data_dir="/data/data/junk")
```
Note that `store_bikedata()` will also download the data if they don't already
exist.

These data can then be accessed with the `RPostgreSQL` package:
```{r, eval=TRUE}
library(RPostgreSQL)
drv <- dbDriver("PostgreSQL")
citibike_con = dbConnect(drv, dbname = "nyc-citibike-data")

query <- function(sql, con = citibike_con) 
  fetch(dbSendQuery(con, sql), n = 1e8)
get_ntrips <- function ()
{
    n <- query("SELECT * FROM station_to_station_counts")
    stns <- sort (unique (n$start_station_id))
    # transpose so index is [from, ti]
    n <- t (array (n$count, dim=rep (length (stns), 2)))
    rownames (n) <- colnames (n) <- stns
    return (n)
}
nt <- get_ntrips ()
junk <- dbDisconnect (citibike_con)
cat (format (sum (nt), big.mark=",", scientific=FALSE), 
     "trips between", dim (nt)[1], "stations\n")
```
(These numbers reflect only a select few of all possible files---there are far
more data than that in total.)

At present the `postgres` database has to be manually removed with
```{r, eval=FALSE}
system ("dropdb nyc-citibike-data")
```

### Test Results

```{r, eval=FALSE}
library(bikedata)
library(testthat)

date()

test_dir("tests/")
```
