---
title: "bikedata"
output:
  md_document:
    variant: markdown_github
    toc: true
  html_document:
    toc: true
    theme: flatly
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Build Status](https://travis-ci.org/mpadge/bikedata.svg)](https://travis-ci.org/mpadge/bikedata) 
[![Project Status: Concept - Minimal or no implementation has been done yet.](http://www.repostatus.org/badges/0.1.0/concept.svg)](http://www.repostatus.org/#concept)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/bikedata)](http://cran.r-project.org/web/packages/bikedata) 


R package to load data from public bicycle hire systems. Currently a
proof-of-concept that only loads data from the New York City [citibike
scheme](https://www.citibikenyc.com/).


# 1. Installation

```{r, eval=FALSE}
devtools::install_github("mpadge/bikedata")
```
```{r usage2, echo=FALSE, message=FALSE}
devtools::load_all (".")
#devtools::load_all (".", recompile=TRUE)
#devtools::document (".")
#goodpractice::gp ("bikedata")
#devtools::check (".")
#testthat::test_dir ("./tests/")
#Rcpp::compileAttributes()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
options(width=120)
```

# 2. Usage

```{r, eval=FALSE, echo=FALSE}
library(bikedata)

# current verison
packageVersion("bikedata")
```

Data from the NYC citibike system can can be downloaded with
```{r, eval=FALSE}
dl_bikedata (data_dir='/tmp')
```
The resultant files can then be read into an `spatialite` database with
```{r, eval=FALSE}
store_bikedata_spl (data_dir='/tmp', 'spdb')
```
```{r, eval=FALSE, echo=FALSE}
store_bikedata_spl (data_dir='/data/data/bikes/nyc-temp', 'spdb')
```

Numbers of trips between each pair of stations ('trip matrices') can then be
obtained in square matrix form from
```{r, eval=FALSE}
tmat <- tripmat ('spdb')
format (sum (tmat), big.mark=',')
```
```{r, echo=FALSE}
"2,886,218"
```
(This is only a small selection of all data.)
```{r, eval=FALSE}
dim (tmat)
```
```{r, echo=FALSE}
c (330, 330)
```

## 2.1. Filtering trips by date

Trip matrices can be constructed for trips filtered by dates and/or times. The
temporal extent of the database can be readily viewed with
```{r, eval=FALSE}
get_datelimits ('spdb')
```
```{r, echo=FALSE}
res <- c ("2014-07-01 00:00:04", "2014-11-16 08:50:59")
names (res) <- c ("first", "last")
res
```
The `tripmat` function accepts four optional arguments specifying `start_date`,
`end_date`, `start_time`, and `end_time`.

A trip matrix constructed from trips beginning after a given date with the
`start_date` argument.
```{r, eval=FALSE}
tmat <- tripmat ('spdb', start_date="20140901")
format (sum (tmat), big.mark=',')
```
```{r, echo=FALSE}
"953,887"
```
As expected, that reduces numbers of trips. End dates can also be specified
```{r, eval=FALSE}
tmat <- tripmat ('spdb', start_date=20140901, end_date="14/09/03")
format (sum (tmat), big.mark=',')
```
```{r, echo=FALSE}
"89,508"
```
Note that dates can be specified in almost any format, as long as the order is
`year-month-day`.

## 2.2. Filtering trips by time of day

Trips can also be selected starting and/or ending at specific times of day.
```{r, eval=FALSE}
tmat <- tripmat ('spdb', start_time=0, end_time=1)
format (sum (tmat), big.mark=',')
```
```{r, echo=FALSE}
"34,512"
```
Single values are interpeted to specify hours, so the above request returns only
those rides start at or after midnight (`00:00:00`) and finishing prior to and
including 1am (`01:00:00`). Single numeric values of `23` are interpreted as the
end of a day (`23:59:59`).

Dates and times can of course be combined
```{r, eval=FALSE}
tmat <- tripmat ('spdb', start_date=20140901, end_date="14,09,03",
                 start_time=6, end_time=9)
format (sum (tmat), big.mark=',')
```
```{r, echo=FALSE}
"10,715"
```

## 2.3. Filtering trips by day of week

Trips can also extracted for particular days of the week by specifying the
`weekday` argument of `tripmat`. Weekdays can be numeric, start from `1=Sunday`,
or any unambiguous character string.
```{r, eval=FALSE}
tmat <- tripmat ('spdb', weekday=7)
format (sum (tmat), big.mark=',')
```
```{r, echo=FALSE}
"350,986"
```

Several weekdays can also be selected
```{r, eval=FALSE}
tmat <- tripmat ('spdb', weekday=c('m', 'Th'))
format (sum (tmat), big.mark=',')
```
```{r, echo=FALSE}
"862,119"
```

Weekdays can also be specified along with other dates and times
```{r, eval=FALSE}
tmat <- tripmat ('spdb', weekday=2:6, start_time=6, end_time=9)
format (sum (tmat), big.mark=',')
```
```{r, echo=FALSE}
"329,919"
```
