---
title: "1. bikedata"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        theme: flatly
vignette: >
  %\VignetteIndexEntry{1. osmdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## 1. Introduction

`bikedata` is an R package for downloading and aggregating data from 
public bicycle hire, or bike share, systems. Although there are very many
public bicycle hire systems in the world ([see this wikipedia
list](https://en.wikipedia.org/wiki/List_of_bicycle-sharing_systems)),
relatively few openly publish data on system usage. The `bikedata` package aims
to enable ready importing of data from all systems which provide it, and will
be expanded on an ongoing basis as more systems publish open data. Cities and
names of associated public bicycle hire systems currently included in the
`bikedata` package, along with numbers of bikes and of docking stations, are:

City | Hire Bicycle System | Number of Bicycles | Number of Docking Stations
--- | --- | --- | ---
London, U.K. | [Santander Cycles](https://tfl.gov.uk/modes/cycling/santander-cycles) | 13,600 | 839
New York City NY, U.S.A. | [citibike](https://www.citibikenyc.com/) | 7,000 | 458
Chicago IL, U.S.A. | [Divvy](https://www.divvybikes.com/) | 5,837 | 576
Washingon DC, U.S.A. | [Capital BikeShare](https://www.capitalbikeshare.com/) | 4,457 | 406
Boston MA, U.S.A. | [Hubway](https://www.thehubway.com/) | 1,461 | 158
Los Angeles CA, U.S.A. | [Metro](https://bikeshare.metro.net/) | 1,000 | 65

All of these systems record and disseminate individual trip data, minimally
including the times and places at which every trip starts and ends. Some provide
additional anonymised individual data, typically including whether or not a user
is registered with the system and if so, additional data including age, gender,
and residential postal code.

Cities with extensively developed systems and cultures of public hire bicycles,
yet which do not provide (publicly available) data include:

City | Number of Bicycles | Number of Docking Stations
--- | --- | ---
Hangzhou, China | 78,000 | 2,965
Paris, France | 14,500 | 1,229
Barcelona, Spain | 6,000 | 424

The current version of the `bikedata` R package can be installed with the
following command:
```{r install, eval = FALSE}
install.packages ('bikedata')
```

```{r, echo = FALSE, message = FALSE, results = 'hide'}
devtools::load_all('..', export_all = FALSE)
```

Once installed, it can be loaded in the usual way:

```{r, eval = FALSE}
library(bikedata)
```

## 2. Main Functions

The `bikedata` function `store_bikedata()` downloads individual trip data from
any or all or the above listed systems and stores them in an `SQLite3` database.
For example, the following line will download all data from the Metro
system of Los Angeles CA, U.S.A., and store them in a database named 'bikedb',
```{r store-la-data, eval=FALSE}
store_bikedata (city = 'la', bikedb = 'bikedb', quiet = TRUE)
```
```{r}
# 131336
```
The function returns the number of trips added to the database.  Both the
downloaded data and the `SQLite3` database are stored by default in the
temporary directory of the current `R` session. The downloaded data are deleted
after having been loaded into the `SQLite3` database, and the database itself
is deleted on termination of the `R` session.  (All of these options may be
overridden as described below.)

Successive calls to `store_bikedata()` will append additional data to the same
database. For example, the following line will append all data from Chicago's
Divvy bike system from the year 2017 to the database created with the first
call above. 
```{r, eval = FALSE}
store_bikedata (bikedb = 'bikedb', city = 'divvy', dates = 2017)
```
The function again returns the number of trips *added* to the database, which
is now less than the total number of trips stored of:
```{r, eval=FALSE}
bike_total_trips (bikedb = 'bikedb')
```

Having stored individual trip data in a database, the primary function of the
`bikedata` package is `bike_tripmat()`, which extracts aggregate numbers of
trips between all pairs of stations. The minimal argument to this function is
the name of the database, along with the name of a city for databases holding
data from multiple cities.
```{r, eval = FALSE}
tm <- bike_tripmat (bikedb = 'bikedb', city = 'la')
dim (tm); sum (tm)
```
The Los Angeles Metro system has 65 docking stations, and there were a total of
XXX individual trips up to April 2017.  Details of the docking stations
associated with these trip matrices can be obtained with
```{r, eval=FALSE}
bike_stations (bikedb = 'bikedb')
bike_stations (bikedb = 'bikedb', city = 'la')
```
For consistency and to avoid potential confusion of function names, most
functions in the `bikedata` package begin with the prefix `bike_` (except for
`store_bikedata()` and `dl_bikedata()`).

Databases generated by the `bikedata` package will generally be very large (at
least several GB), and many functions may take considerable time to execute.
It is nevertheless possible to explore package functionality *quickly* through
using the additional helper function, `bike_write_test_data()`. This function
uses the `bike_dat` data set provided with the package, which contains details
of 200 representative trips for each of the cities listed above. The function
writes these data to disk as `.zip` files which can then be read by the
`store_bikedata()` function.
```{r, eval=FALSE}
bike_write_test_data ()
store_bikedata (bikedb = 'testdb')
bike_summary_stats (bikedb = 'testdb')
```
The `.zip` files generated by `bike_write_test_data()` are created by default
in the `tempdir()` of the current `R` session, and so will be deleted on
session termination. Specifying any alternative `bike_dir` will create enduring
copies of those files in that location which ought to be deleted when finished.

The remainder of this vignette provides further detail on these three distinct
functional aspects of downloading, storage, and extraction of data.

## 3. Downloading Data

The `store_bikedata()` function demonstrated above automatically downloads data
and deletes the downloaded files once the data has been loaded into the
`SQLite3` database. Enduring copies of the raw data files may be created with
the function `dl_bikedata()`, and specifying a (non-default) location, such as,
```{r, eval=FALSE}
dl_bikedata (city = 'chicago', data_dir = '/data/bikedata/')
```

## 3.1 Downloading data for specific date ranges

Both `store_bikedata()` and `dl_bikedata()` accept an additional argument
(`dates`) specifying ranges of dates for which data should be downloaded and
stored.  The format of this argument is quite flexible so that,
```{r, eval = FALSE}
dl_bikedata (city = 'dc', dates = 16)
```
will download data from Washington DC's Capital Bikeshare system for all 12
months of the year 2016, while,
```{r, eval = FALSE}
dl_bikedata (city = 'ny', dates = 201604:201608)
```
will download New York City data from April to August (inclusively) for that
year. (Note that the default `data_dir` is the `tempdir()` of the current `R`
session, with downloaded files being deleted upon session termination.) Dates
can also be entered as character strings, with the following calls producing
results equivalent to the preceding call, 
```{r, eval = FALSE}
dl_bikedata (city = 'ny', dates = '2016/04:2016/08')
dl_bikedata (city = 'new york', dates = '201604:201608')
dl_bikedata (city = 'n.y.c.', dates = '2016-04:2016-08')
dl_bikedata (city = 'new', dates = c (201604, 201605, 201606, 201607, 201608))
dl_bikedata (city = 'CI', dates = 201600 + 4:8)
```
The arguments `city = 'new'` and `city = 'CI'` in the final call are sufficient
to uniquely identify New York City's citibike system.

If files have been previously downloaded to a nominated directory, then calling
the `dl_bikedata()` function will only download those data files that do not
already exist. This function may thus be used to periodically refresh the
contents of a nominated directory as new data files become available.

## 4. Storing Data

As mentioned above, individual trip data are stored in a single `SQLite3`
database, created by default in the temporary directory of the current `R`
session. Specifying a path for the `bikedb` argument in the `store_bikedata()`
function will create a database that will remain in that location until
explicitly deleted. 

The nominated database is created if it does not already exist, otherwise
additional data are appended to the existing database. As described above, the
same `dates` argument can be passed to both `dl_bikedata()` and
`store_bikedata()` to download data within specified ranges of dates.

Both `dl_bikedata()` and `store_bikedata()` are primarily intended to be used to
download data for specified cities. Calling either function without specifying a
city will simply request confirmation from the user that data from *all* cities
really ought to be downloaded and/or stored. 

```{r, eval = FALSE}
ntrips <- dl_bikedata (bikedb = 'bikedb')
ntrips <- store_bikedata (bikedb = 'bikedb')
```

### 4.1 Storing and loading enduring copies of trip data files

Enduring copies of trip files may be created with the `dl_bikedata()` function
described above. These pre-downloaded files may also be passed directly to
`store_bikedata()` by simply specifying the data directory in the `data_dir`
argument. The following call will load all previously downloaded files in the
specified directly in the nominated database.
```{r, eval = FALSE}
store_bikedata (bikedb = 'bikedb', data_dir = '/data/bikedata/')
```
As described above, the function `dl_bikedata()` may be used to periodically
refresh downloaded files when new data become available. The `store_bikedata()`
function provides a similar capability. When called without specifying
`data_dir`, the function will download only those files which have not been
previously stored in the database, whereas when called with a specific
`data_dir`, the function will download any files not present in the nominated
directly and load them in to the database.

In short, the `store_bikedata()` function may be repeatedly called to load only
those data published since the last time the function was called, while
enduring copies of the raw data files on individual trips may be periodically
refreshed with `dl_bikedata()`, and the associated directory specified in the
call to `store_bikedata()` to load only recently added files.


### 4.2 Database Indexes

The `store_bikedata()` function also has a default option to create indexes for
more efficient data extraction. Indexes are always created for cities, with the
`create_index = TRUE` option creating additional indexes for dates and times.
Setting this parameter to `FALSE` will enable faster initial storage of data,
yet is likely to lead to slower aggregation of data based on dates and times
(see below). It is therefore recommended to set this option to `FALSE` only when
subsequent data aggregation will **not** be based on dates or times.

## 5. Accessing Aggregate Data

### 5.1 Origin-Destination Matrices

As briefly described in the introduction, the primary function for extracting
aggregate data from the `SQLite3` database established with `store_bikedata()`
is `bike_tripmat()`. With the single mandatory argument naming the database,
this function returns a matrix of numbers of trips between all pairs of
stations.  Trip matrices can be returned either in square form (the default),
with both rows and columns named after the bicycle docking stations and matrix
entries tallying numbers of rides between each pair of stations, or in long form
by requesting `bike_tripmat (..., long = TRUE)`. The latter case will return a
[`tibble`](https://cran.r-project.org/package=tibble) with the three columns of
`station_station_id`, `end_station_id`, and `number_trips`.
```{r, eval=FALSE}
tm <- bike_tripmat (bikedb = 'bikedb', city = 'la')
head (tm)
```
It is possible to extract a square matrix aggregating trips between all
stations from all cities within the database, yet doing so prompts a message
explaining that it is better to extract trip matrices for particular cities. 
```{r, eval=FALSE}
tm <- bike_tripmat (bikedb = 'bikedb', long = FALSE)
```
Absent city specification, the resultant trip matrix will include trips between
bike stations in different cities, for which there will obviously never be any
recorded trips (even if only because of incompatible docking stations). For two
cities, for example, maximally one half of the resultant matrix can contain
non-zero entries; for three cities, one ninth, and so on. In contrast, single
long-form trip matrices may be aggregated for all cities without generating
such a message, because in this case the matrix includes only those pairs of
stations with one or more trips between them., 
```{r, eval=FALSE}
tm <- bike_tripmat (bikedb = 'bikedb', long = TRUE)
```


### 5.1.1. Temporal filtering of trip matrices

Trip matrices can also be extracted for particular dates, times, and days of
the week, through specifying one or more of the optional arguments:

1. `start_date`
2. `end_date`
3. `start_time`
4. `end_time`
5. `weekday`

Arguments may in all cases be specified in a range of possible formats as long
as they are unambiguous, and as long as 'larger' units precede 'smaller' units
(so years before months before days, and hours before minutes before seconds).
Acceptable formats may be illustrated through specifying a list of arguments to
be passed to `bike_tripmat()`:
```{r, eval=FALSE}
args <- list (start_date = 16, end_time = 12, weekday = 1)
tm <- bike_tripmat (bikedb = 'bikedb', city = 'ny', args)
```
where a two-digit `start_date` (or `end_date`) is interpreted to represent a
year, while a one- or two-digit `_time` is interpreted to represent an hour.  A
value of `end_time = 24` is interpreted as `end_time = '23:59:59'`, while a
value of `_time = 0` is interpreted as `00:00:00`.  The following further
illustrate the variety of acceptable formats,
```{r, eval=FALSE}
args <- list (start_date = '2016 May', end_time = '12:39', weekday = 2:6)
args <- list (end_date = 20160720, end_time = 123915, weekday = c ('mo', 'we'))
args <- list (end_date = '2016-07-20', end_time = '12:39:15', weekday = 2:6)
```
Both `_date` and `_time` arguments may be specified in either `character` or
`numeric` forms; in the former case with arbitrary (or no) separators.
Regardless of format, larger units must precede smaller units as explained
above.

Weekdays may specified as characters, which must simply be unambiguous and (in
admission of currently inadequate internationalisation) correspond to standard
English names. Minimal character specifications are thus `'so', 'm', 'tu', 'w',
'th', 'f', 'sa'`. The value of `weekday = 1` denotes Sunday, so `weekdays =
2:6` denote the traditional working days, Monday to Friday, while weekends may
be denoted with `weekdays = c ('sa', 'so')` or `weekdays = c (1, 7)`.


### 5.2. Standardising trip counts by durations of operation

Most bicycle hire systems have progressively expanded over time through ongoing
addition of new docking stations. Total numbers of counts within a trip matrix
will thus be generally less for more recently installed stations, and more for
older stations. The `bike_tripmat()` function has an option, `standardise =
FALSE`. Setting `standardise = TRUE` (or `standardize`, if you'd rather spell it
that way) allows trip matrices to be standardised for durations of station
operation, so that numbers of trips between any pair of stations reflect what
they would be if all stations had been in operation for the same duration.

Standardisation implements a linear scaling of total numbers of trips to and
from each station according to total durations of operation, with counts in
the final trip matrix scaled to have the same total number of trips as the
original matrix.  This standardisation has two immediate consequences:

1. Numbers of trips will not necessarily be integer values, but are rounded for
the sake of sanity to three digits, corresponding to the maximal likely
precision attainable for daily differences in operating durations;
2. Trip numbers will generally not equal actual observed numbers. Counts for
the longest operating durations will be lower than actually recorded, while
counts for more recent stations will be greater than observed values.

The `standardise` option nevertheless enables travel patterns between different
(groups of) stations to be statistically compared in a way that is free of the
potentially confounding influence of differing durations of operation.


### 5.4. Demographic filtering of trip matrices

As described at the outset, the bicycle hire systems of several cities provide
additional demographic information including whether or not cyclists are
registered with the system, and if so, additional information including birth
years, genders, and postal (zip) codes. Note that the provision of such
information is voluntary, and that no providers can or do guarantee the accuracy
of their data.

Those systems which provide demographic information are listed with the
function `bike_demographic_data()`, which also lists the nominal kinds of
demographic data provided by the different systems. Data can then be filtered
by demographic parameters with additional optional arguments to
`bike_tripmat()` of,

1. `registered` (`TRUE/FALSE`, `'yes'/'no'`, 0/1)
2. `birth_year` (as one or more four-digit numbers or character strings)
3. `gender` ('m/f/.', 'male/female/other')
4. `post_code` (as five-digit number or character string)

Many systems allow users to specify alternative genders, and any values not
beginning with either `'m'` or `'f'` (case-insensitive) are interpreted as
requests for alternative genders. Because there is no systematic way to specify
alternative genders, additional filtering beyond these three classes of gender
is currently not possible.  Note further that, although the only systems to
date which provide postal addresses are from the U.S., this parameter is not
called `zip` to avoid potential association with the `.zip`-format files
generally provided by the hire bicycle systems and accessed by `bikedata`.

It is of course possible to combine all of these optional parameters in a
single query. For example,
```{r, eval=FALSE}
tm <- bike_tripmat (bikedb = 'bikedb', city = 'ny', start_date = 2016,
        start_time = 9, end_time = 24, weekday = 2:6, gender = 'xx', 
        birth_year = 1900:1950)
```
The value of `gender = 'xx'` will be interpreted to request data from all
members with nominal alternative genders.  As demographic data are only given
for registered users, the `registered` parameter is redundant in this query.
Note further than many systems offer a range of potential birth years starting
from a default value of 1900, and there are consequently a significant number of
cyclists who declare this as their birth year.


### 5.5. Station Data

Data on docking stations may be accessed with the function `bike_stations()`
```{r, eval = FALSE}
bike_stations (bikedb = 'bikedb')
```
This function returns a `tibble` detailing the names and locations of all
bicycle stations present in the database. Station data for specific cities may
be extracted through specifying an additional `city` argument.
```{r, eval = FALSE}
bike_stations (bikedb = 'bikedb', city = 'ny')
```


### 5.6. Summary Statistics

`bikedata` provides a number of helper functions for extracting summary
statistics from the `SQLite3` database. The function `bike_summary_stats
(bikedb)` generates an overview table. (This function may take some time to
execute on large databases.)

```{r, eval=FALSE}
bike_summary_stats ('bikedb')
# Total of --- trips from 5 cities
# 
# city  |   first trip  |   last trip   | number of trips
# ------|---------------|---------------|-----------------
# 
#  ...
```

Additional helper functions which provide individual components from this
summary data (and which will generally do so notably faster for large databases
than the above function) include:
```{r, eval=FALSE}
bike_num_trips (bikedb = 'bikedb', city = 'ny')
bike_date_range (bikedb = 'bikedb', city = 'ny')
```


## 6. Direct database access


Although the `bikedata` package aims to circumvent any need to access the
database directly, through providing ready extraction of trip data for most
analytical or visualisation needs, direct access may be achieved either using
the convenient `dplyr` functions, or the more powerful functionality provided
by the `RSQLite` package.

```{r, eval=FALSE}
# demo list of tables and contents of both trips and stations tables
```

## 7. Visualisation of bicycle trips

The `bikedata` package does not provide any functions enabling visualisation of
aggregate trip data, both because of the primary focus on enabling access and
aggregation in the simplest practicable way, and because of the myriad
different ways users of the package are likely to want to visualise the data.
This section therefore relies on other packages to illustrate some of the ways
in which trip matrices may be visualised.

The simplest spatial visualisation involves connecting the geographical
coordinates of stations with straight lines, with numbers of trips represented
by some characteristics of the lines connecting pairs of stations, such as
thickness or colours.  This can be achieved with the following code, which also
illustrates that it is generally more useful for visualisation purposes to
extract trip matrices in long rather than square form.
```{r, eval=FALSE}
stns <- bike_stations (bikedb = 'bikedb', city = 'nyc')
ntrips <- bike_tripmat (bikedb = 'bikedb', city = 'nyc', long = TRUE)
x1 <- stns$longitude [match (stns$longitude, ntrips$start_station)]
y1 <- stns$latitude [match (stns$longitude, ntrips$start_station)]
x2 <- stns$longitude [match (stns$longitude, ntrips$end_station)]
y2 <- stns$latitude [match (stns$longitude, ntrips$end_station)]
plot (stations$longitude, stations$latitude)
lines (cbind (x1, y1), cbind (x2, y2), 
        lwd = ntrips$num_trips * 10 / max (ntrips$num_trips))
```
