---
title: "1. bikedata"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        theme: flatly
vignette: >
  %\VignetteIndexEntry{1. osmdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## 1. Introduction

`bikedata` is an R package for downloading and aggregating data from public
bicycle hire systems. Cities and the names of associated public bicycle Systems
currently included are:

1. New York citibike
2. Boston Hubway
3. Chicago Divvy
4. Washingon D.C. Capital BikeShare
5. Los Angeles Metro

`bikedata` also facilitates access to data from the Barclay's Bicycle Hire
system of London, U.K., for which users must first register to obtain an
`OAuth` key to access the data.

Cities with bike systems which do not currently provide data, but for which
`bikedata` will be adapted as soon as data become available include:

1. Miami, Florida
2. Portland, Oregon

The development version of the `bikedata` R package can be installed with the
`devtools` package using the following command:

```{r install, eval=FALSE}
devtools::install_github('mpadge/bikedata')
```

```{r, echo=FALSE, message=FALSE, results='hide'}
devtools::load_all('..', export_all = FALSE)
```

Once installed, it can be loaded in the usual way:

```{r, eval=FALSE}
library(bikedata)
```

## 2. Main Functions

All of the above listed public bicycle hire systems record and disseminate
individual trip data, minimally including the times and places at which every
trip starts and ends. Some systems provide additional anonymised individual
data, typically including whether or not a user is registered with the system
and if so, the age, gender, and residential postal code.

Data are downloaded with the single function, `dl_bikedata()`,
```{r, eval=FALSE}
dl_bikedata (city="chicago")
```
This function downloads the requested data (in this case, all data from
Chicago's Divvy Bike system) and stores them in the temporary directory of the
current `R` session (although this location may be overridden). The
`dl_bikedata()` function will only download those data files that do not
already exist, and thus function may thus be used to periodically refresh the
contents of a nominated directory.

Having downloaded data, they can then be collected stored in a single `SQLite3`
database with the single call 
```{r, eval=FALSE}
ntrips <- store_bikedata (data_dir=tempdir(), bikedb = "bikedb")
```
This will load the trip data stored in individual files in the location
`data_dir` into the nominated database. A database will be created if none
exists, otherwise data will be appended to existing data. Cities associated
with data files held in the nominated location are identified automatically,
because each system has its own unique way of naming files. The function
returns the total number of trips added to the database.

Finally, an ''origin-destination'' matrix of numbers of trips between all pairs
of stations can be aggregated with the call, 
```{r, eval=FALSE}
tm <- tripmat (bikedb = "bikedb")
```
while details of the stations can be obtained with
```{r, eval=FALSE}
bike_stations (bikedb = "bikedb")
```

The remainder of this vignette details these three distinct functional aspects
- downloading, storage, and extraction of data.

## 3. Downloading Data

The function `dl_bikedata()` requires only the single argument of the city or
bicycle hire system for which data are desired, and accepts the two additional
optional arguments of `data_dir` and `dates`. The latter of these can be used
to download only those files within a nominated range of calendrical dates, and
must be passed as a vector listing all desired data. For example,
```{r, eval=FALSE}
dl_bikedata (city='dc', dates=c('2015.11', '2015.12', 'Jan 2016'))
```
It is not necessary to explicitly download data with the `dl_bikedata()`
function - the function `store_bikedata()` described in the following section
will automatically download requested data absent specification of a local
storage location. The `dl_bikedata()` function is primarily intended to allow
direct user-control over the storage of raw trip data files.

As mentioned, data from the Barclay's Bicycle Hire Scheme of London, U.K.
require registration with [Transport for London](tfl.gov.uk). Registration
enables access to an `OAuth` key which must additionally be passed to the
`dl_bikedata()` function to access these data.

## 4. Storing Data

As mentioned above, individual trip data are stored in a single `SQLite3`
database. Specifying any path other than the default `data_dir = tempdir()` in
the `store_bikedata()` function call will create a database that will remain in
that location until explicitly deleted. 

A database may be initiated with an initial call to `store_bikedata()`:
```{r, eval=FALSE}
ntrips <- store_bikedata (bikedb = "bikedb", city="chi")
```
Although the `bikedata` package aims to circumvent any need to access the
database directly, through providing ready extraction of trip data for most
analytical needs, direct access may be achieved either using the convenient
`dplyr` functions, or the more powerful functionality provided by the `RSQLite`
package.

```{r, eval=FALSE}
# demo list of tables and contents of both trips and stations tables
```

This reveals that data are stored in two primary tables, one for all individual
trips and one detailing the locations and other properties of the bicycle
docking stations.

The parameter `data_dir` of the `store_bikedata()` function specifies the
location where previously downloaded trip data are stored. If not given,
requested data are downloaded into `tempdir()` and deleted after they have been
loaded into the database.

In addition to naming the database for storing trip data, `store_bikedata()`
minimally requires naming the city for which to download data.
```{r, eval=FALSE}
store_bikedata (bikedb = 'bikedb', city = 'dc', dates = 2016)
```
Loading data from other cities requires repeated calls to `store_bikedata()`.
The entirety of currently available data can thus be loaded into a single
database with the following lines:
```{r, eval=FALSE}
cities <- bike_cities ()
for (c in cities)
    store_bikedata (bikedb = 'bikedb', city = c)
```
`bikedata` is explicitly designed *not* to permit automatic downloading of all
data in a single call, lest an innocent exploratory usage inadvertently
generate a database 5GB in size.

The `store_bikedata()` function also enables the nominated database to be
updated to reflect more recently published data.

This database may be updated as new data become available (generally on a
monthly basis).

